Building and Checking Suffix Array Using Induced Sorting Method

### 2017-04-21 new-draft/draft.tex
2017/04/21

\documentclass[10pt,journal,compsoc]{IEEEtran}
\usepackage[nocompress]{cite}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{url}
\usepackage{multirow}
\usepackage[ruled]{algorithm2e}
\usepackage{enumitem}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{observation}[theorem]{Observation}

@@

\title{Efficient Suffix Array Construction and Checking in External Memory}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%
%
%\IEEEcompsocitemizethanks is a special \thanks that produces the bulleted
% lists the Computer Society journals use for "first footnote" author
% affiliations. Use \IEEEcompsocthanksitem which works much like \item
% for each affiliation group. When not in compsoc mode,
% \IEEEcompsocitemizethanks becomes like \thanks and
% \IEEEcompsocthanksitem becomes a line break with idention. This
% facilitates dual compilation, although admittedly the differences in the
% desired content of \author between the different types of papers makes a
% one-size-fits-all approach a daunting prospect. For instance, compsoc 
% journal papers have the author affiliations above the "Manuscript
% received ..."  text while in non-compsoc journals this is reversed. Sigh.

@@

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
%\markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
%{Shell \MakeLowercase{\textit{et al.}}: Bare Advanced Demo of IEEEtran.cls for IEEE Computer Society Journals}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% The publisher's ID mark at the bottom of the page is less important with
% Computer Society journal papers as those publications place the marks
% outside of the main text columns and, therefore, unlike regular IEEE
% journals, the available text space is not reduced by their presence.
% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2015 IEEE}
% or like this to get the Computer Society new two part style.
%\IEEEpubid{\makebox[\columnwidth]{\hfill 0000--0000/00/\$00.00~\copyright~2015 IEEE}%
%\hspace{\columnsep}\makebox[\columnwidth]{Published by the IEEE Computer Society\hfill}}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark (Computer Society journal
% papers don't need this extra clearance.)

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEtitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\IEEEtitleabstractindextext{%
\begin{abstract}

A suffix array (SA) must be checked to ensure its correctness. The existing SA checker verifies an SA by performing two passes of integer sorts that are time and space consuming especially when running in the external memory. Currently, building and checking an SA are done in sequence, where the time and the space bottlenecks are caused by building and checking the SA, respectively.
In this paper, a fast and lightweight checking method is proposed to enable any induced sorting (IS) suffix sorting algorithm to verify a suffix array while it is being built. For performance evaluation, we first redesign the reduction phase of the external memory IS algorithm DSA-IS by employing the new substring sorting and naming methods to produce an optimized alternative DSA-IS+, and then apply the new checking method to DSA-IS+ for building and checking an SA simultaneously. Our experimental results indicate that, in comparison with that for building, the checking time and space complexities of eSAIS are substantial, but that of DSA-IS+ are negligible. As a result, DSA-IS+ can build and check an SA in a higher total time and space efficiencies.

@@

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
suffix array, construction and verification, external memory.
\end{IEEEkeywords}}

@@

% To allow for easy dual compilation without having to reenter the
% abstract/keywords data, the \IEEEtitleabstractindextext text will
% not be used in maketitle, but will appear (i.e., to be "transported")
% here as \IEEEdisplaynontitleabstractindextext when compsoc mode
% is not selected <OR> if conference mode is selected - because compsoc
% conference papers position the abstract like regular (non-compsoc)
% papers do!
\IEEEdisplaynontitleabstractindextext
% \IEEEdisplaynontitleabstractindextext has no effect when using
% compsoc under a non-conference mode.

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\IEEEraisesectionheading{\section{Introduction}\label{sec:introduction}}

\subsection{Background}
\IEEEPARstart{T}{he} suffix array~(SA)~\cite{Manber1993} is a data structure that has been widely used in many string processing applications, e.g., biological sequence alignment, time series analysis and text clustering. Given an input string, traversing its suffix tree can be emulated by using the corresponding enhanced suffix array~\cite{Abouelhodaa2004}, which mainly consists of the suffix and the longest common prefix arrays. It has been realized that the application scope of an index mainly depends on the construction speed and the space requirement. This leads to intensive works on designing time and space efficient suffix sorting algorithms over the past decade, assuming different computation models such as internal memory, external memory, parallel and distributed models. Particularly, to keep pace with the fast development of the data sampling techniques, several external memory algorithms have been proposed for building massive suffix arrays in recent years, e.g., DC3~\cite{Dementiev2008a}, bwt-disk~\cite{Ferragina2012}, SAscan~\cite{Karkkainen2014}, pSAscan~\cite{Karkkainen2015}, eSAIS~\cite{Bingmann2013}, EM-SA-DS~\cite{Nong2014} and DSA-IS~\cite{Nong2015}. Among them, the latter three algorithms are based on the induced sorting~(IS) method described in SA-IS~\cite{Nong2011}.

The basic idea behind the induced sorting method is to induce the lexicographical order of all the substrings/suffixes from a sorted subset of substrings/suffixes. Following the idea, an IS-based suffix sorting algorithm is typically comprised of a reduction phase for sorting and naming substrings to reduce a string $x[0,n)$ to a short string $x1[0,n1)$ with $n1 \le \frac{1}{2}n$ and an induction phase for sorting suffixes to induce ${\sf SA}(x)$ from ${\sf SA}(x1)$. During the two phases, the key operation is to retrieve the preceding character of a sorted substring/suffix. This can be done very quickly when $x$ is fully accommodated in the internal memory, but will become slow when $x$ resides in the external memory, as each operation takes a random disk access. For a high I/O efficiency, eSAIS, EM-SA-DS and DSA-IS use different auxiliary data structures to retrieve the preceding characters in a disk-friendly way. Particularly, both eSAIS and EM-SA-DS split a long substring into pieces and represent each piece by a fixed-size tuple, while DSA-IS does not. With an elaborate arrangement of the I/O operations, the programs for these three algorithms are competitive with those for others in terms of both time and space efficiencies.

\subsection{Problems to Solve}
Although the above three IS-based algorithms achieve great performance in comparison with prior arts, there still exist several drawbacks to be dealt with.

Firstly, the space requirement remains at a high level for all the three algorithms. In details, the peak disk usages for eSAIS, EM-SA-DS and DSA-IS are around $24n$, $31n$ and $20n$, respectively, the minimum of which is about three times as that for the lightweight suffix sorter SAscan (which I/O volume is super-linear). It is identified that DSA-IS reaches the peak disk usage when induced sorting substrings during the reduction phase. This reveals a need for designing new substring sorting and naming methods to improve the space efficiency in order to scale the problem size that can be tackled.

Secondly, a suffix array must be checked to ensure its correctness for later use. Currently, an SA built by each of these algorithms is verified using the SA checker described in~\cite{Karkkainen2003}, which performs two passes of external memory integer sorts that charge large disk space and long cpu time. As a result, the peak disk usage for both eSAIS and DSA-IS rises up to $26n$ while the time consumption also grows by 20 percent as before, leading to a substantial performance degradation especially when processing large-scale datasets. Therefore, it is desired to reduce the check overhead in case it becomes a performance bottleneck.

\subsection{Our Contribution}

Our contribution in this paper mainly includes the following aspects.

\begin{itemize}

	\item We redesign the reduction phase of DSA-IS by employing new methods for sorting and naming substrings. In our experiments, the program for the enhanced algorithm, called DSA-IS+, runs as fast as that for eSAIS while its peak disk usage is about two-thirds as that of eSAIS.

	\item We design a fast and lightweight SA checking method that can be seamlessly integrated into any IS-based suffix sorting algorithm. For performance evaluation, we combine this method with DSA-IS+ to produce a unified solution that can be employed to construct and check SA at the same time. It was observed that, by exploiting the use of our checking method, the performance overhead for verification is considerable small in comparison with that for construction. In details, the peak disk usage for the unified solution remains at the same level of that for DSA-IS+, while the growth in both time and I/O volume is negligible.

\end{itemize}

The rest of this paper is organized as follows. Section~\ref{sec:preliminary} gives the preliminaries in our presentation. Section~\ref{sec:dsais_plus} describes the substring sorting and naming methods for DSA-IS+, Section~\ref{sec:sachecker} the checking method and Section~\ref{sec:experiments} the experimental results. Finally, we come to the conclusion in Section~\ref{sec:conclusion}.

\section{Preliminaries}\label{sec:preliminary}

\subsection{Basic Notations}

Given an input string $x[0,n)=x[0]x[1]...x[n - 1]$ drawn from a totally ordered alphabet $\delta$, we assume $x[n - 1]$ to be the smallest character in $\delta$ that appears in $x$ only once. For short, we use ${\sf sub}(x,i)$ and ${\sf suf}(x,i)$ to denote a substring running from $x[i]$ to the leftmost LMS character on its right side and a suffix running from $x[i]$ to $x[n - 1]$, respectively. To follow lists the basic notations used in our presentation.

\begin{itemize}[itemindent = 0ex]
	\item L-type, S-type and LMS character/substring/suffix. We say $x[i]$ is S-type if (1) $i = n - 1$ or (2) $x[i] < x[i + 1]$ or (3) $x[i] = x[i+1]$ and $x[i+1]$ is S-type; otherwise, $x[i]$ is L-type. Furthermore, if $x[i]$ and $x[i - 1]$ are respectively S-type and L-type, then $x[i]$ is also an LMS character. Moreover, if $x[i]$ is L-type, S-type or LMS, then ${\sf sub}(x,i)$ and ${\sf suf}(x,i)$ are L-type, S-type or LMS, respectively.

	\item preceding and succeeding character/substring/suffix. We say $x[i - 1]$, ${\sf sub}(x, i - 1)$ and ${\sf suf}(x, i - 1)$ are the preceding character, substring and suffix for $x[i]$, ${\sf sub}(x, i)$ and ${\sf suf}(x, i)$, respectively. Similarly, we say $x[i + 1]$, ${\sf sub}(x, i + 1)$ and ${\sf suf}(x, i + 1)$ are the succeeding character, substring and suffix for $x[i]$, ${\sf sub}(x, i)$ and ${\sf suf}(x, i)$, respectively.

	\item ${\sf SA}$ and ${\sf STRA}$. We respectively use ${\sf SA}$ and ${\sf STRA}$ to denote the suffix and substring arrays. Specifically, both ${\sf SA}(x)$ and ${\sf STRA}(x)$ are permutations of $[0, n)$ such that ${\sf suf}(x, {\sf SA}(x)[i]) < {\sf suf}(x, {\sf SA}(x)[i + 1])$ and ${\sf sub}(x, {\sf SA}(x)[i]) \le {\sf sub}(x, {\sf SA}(x)[i + 1])$ in lexicographical order for all $i \in [0, n - 1)$.

	\item ${\sf BKT}$. Suffixes in ${\sf SA}(x)$ are naturally grouped into buckets according to their head characters. Given ${\sf BKT}({\sf SA}(x), ch)$, the bucket occupies a consecutive range in ${\sf SA}(x)$ that gathers all the suffixes with a head character $ch$. Furthermore, a bucket, say ${\sf BKT}({\sf SA}(x), ch)$, can be subdivided into ${\sf BKT_L}({\sf SA}(x), ch)$ and ${\sf BKT_S}({\sf SA}(x), ch)$ that only contain L-type and S-type suffixes in ${\sf BKT}({\sf SA}(x), ch)$, respectively. Similarly, we define ${\sf BKT}({\sf STRA}(x), ch)$, ${\sf BKT_L}({\sf STRA}(x), ch)$ and ${\sf BKT_S}({\sf STRA}(x), ch)$ for ${\sf STRA}(x)$.

	\item ${\sf ISA}$. Inverse suffix array, satisfying ${\sf ISA}(x)[{\sf SA}(x)[i]] = i$ for all $i \in [0,n)$.

	\item ${\sf PA}$. Position array, where ${\sf PA}(x)[i]$ records the position index of the $i$-th LMS substring from the left side of $x$.

\end{itemize}

\subsection{An Overview of DSA-IS}

As with other IS-based suffix sorting algorithms, DSA-IS mainly consists of a reduction phase followed by an induction phase. The induced sorting method is adopted in both two phases to sort substrings and suffixes, respectively.

\subsubsection{Reduction Phase}

Assume $x[0,n)$ and ${\sf STRA}(x)$ can be wholly accommodated into RAM, SA-IS performs the following three steps in the reduction phase to determine the lexicographical order of substrings.

\begin{itemize}[itemindent = 5ex]
	\item [step 1:] Scan $x$ rightward with $i$ decreasing from $n - 1$ to $0$. For each scanned LMS character $x[i]$, insert $i$ into the rightmost empty position of ${\sf BKT_S}({\sf STRA}(x), ch)$.

	\item [step 2:] Scan ${\sf STRA}(x)$ leftward with $i$ increasing from $0$ to $n - 1$. For each scanned $p = {\sf STRA}(x)[i]$, if $ch = x[p - 1]$ is the starting character of an L-type substring, then insert $p - 1$ into the leftmost empty position of ${\sf BKT_L}({\sf STRA}(x), ch)$.

	\item [step 3:] Scan ${\sf STRA}(x)$ rightward with $i$ decreasing from $n- 1$ to $0$. For each scanned $p = {\sf STRA}(x)[i]$, if $ch = x[p - 1]$ is the starting character of an S-type substring, then insert $p - 1$ into the rightmost empty position of ${\sf BKT_S}({\sf STRA}(x), ch)$.
\end{itemize}

The key operation of steps 2-3 is to retrieve the preceding character $x[p - 1]$ for currently scanned substring starting at $x[{\sf STRA}(x)[i]]$, where each operation takes a random access to the internal memory. A naive solution for extending the above 3-step procedure to external memory models is to replace each memory access with a disk access. However, this is not practical due to the large overhead for $\mathcal{O}(n)$ disk accesses. To reduce the I/O complexity, DSA-IS introduces the alternative ${\sf DSTRA}(x)$ for ${\sf STRA}(x)$ and employs I/O buffers to cache ${\sf DSTRA}(x)$ and other data structures for amortizing the disk access overhead, where the involved data structures are defined as below.

\begin{itemize}[itemindent = 0ex]

	\item ${\sf DSTRA}$, ${\sf DSTRA_L}$ and ${\sf DSTRA_S}$. DSAITEM arrays. For an input $x$, each item of ${\sf DSTRA}(x)$ associates with a substring ${\sf sub}(x, p)$ and mainly consists of the following three components:
	\begin{itemize}[itemindent = 0ex]
		\item $p$: position index for ${\sf sub}(x, p)$.
		\item $c$: head character for ${\sf sub}(x, p)$, that is $x[p]$.
		\item $t$: type of the preceding character $x[p - 1]$, set as 0 or 1 for L-type or S-type, respectively.
	\end{itemize}

	${\sf DSTRA_L}(x)$ and ${\sf DSTRA_S}(x)$ only contain the items of L-type and S-type substrings in ${\sf DSTRA}(x)$, respectively.

	\item ${\sf LMSDATA}$. A string array. Each item stores the characters of an LMS substring.

	\item ${\sf LMSNAME}$. An integer array. Each item stores the name of an LMS substring, where the name represents the substring's lexicographical order among all.

\end{itemize}

Notice that the above data structures organize the items in the external memory according to the lexicographical order of their corresponding substrings.

% Algorithm
\begin{algorithm}
	\SetAlgoNoLine
	\KwIn{$x$}
	\KwOut{$x1$}

	step 1: Partition $x$ into blocks $b_1, b_2, ..., b_k$.~\label{alg:dsais_reduction_phase:1} \\
	step 2: Induced sort the substrings of $b_i$ for all $i \in [1, k]$ to compute ${\sf DSTRA_L}(b_i)$, ${\sf DSTRA_S}(b_i)$ and ${\sf LMSDATA}(b_i)$. ~\label{alg:dsais_reduction_phase:2}\\
	step 3: Induced sort the substrings of $x$ from ${\sf DSTRA_L}(b_i)$, ${\sf DSTRA_S}(b_i)$ and ${\sf LMSDATA}(b_i)$ to compute ${\sf DSTRA}(x)$ and ${\sf LMSDATA}(x)$. Meanwhile, scan ${\sf DSTRA}(x)$ to compute ${\sf DSTRA_{LMS}}(x)$.~\label{alg:dsais_reduction_phase:3}\\
	step 4: Scan ${\sf DSTRA_{LMS}}(x)$ and ${\sf LMSDATA}(x)$ to name the sorted LMS substrings for producing ${\sf LMSNAME}(x)$. ~\label{alg:dsais_reduction_phase:4}\\
	step 5: Compute the reduced string $x1$ from ${\sf LMSNAME}(x)$ and ${\sf DSTRA_{LMS}}(x)$. ~\label{alg:dsais_reduction_phase:5}\\

	\caption{The Reduction Phase for DSA-IS}
	\label{alg:dsais_reduction_phase}
\end{algorithm}

The algorithmic framework of the reduction phase for DSA-IS is shown in Algorithm~\ref{alg:dsais_reduction_phase}, where each step is sketched below.

>>>>

% !TeX spellcheck = en_US

\documentclass[10pt,journal,compsoc]{IEEEtran}
\usepackage{graphicx}
\usepackage[ruled, linesnumbered]{algorithm2e}
\usepackage{url}
\usepackage{epstopdf}
\usepackage{indentfirst}
\usepackage[tight,footnotesize]{subfigure}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{multirow}
\usepackage{color}
\usepackage{enumerate}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{observation}[theorem]{Observation}
\newtheorem{corollary}[theorem]{Corollary}

% *** CITATION PACKAGES ***
\ifCLASSOPTIONcompsoc
\usepackage[nocompress]{cite}
\else
\usepackage{cite}
\fi

@@

\title{Building and Checking Suffix Array Using Induced Sorting Method}

@@

\IEEEtitleabstractindextext{% 
\begin{abstract}

In the past decade, the induced sorting~(IS) method has been successively used to design disk-based suffix array~(SA) construction algorithms that have better time and I/O complexities than the existing alternatives. However, their current programs commonly suffer from a performance bottleneck due to large disk space in need. Recently, it was reported a carefully engineered IS suffix sorter that takes less than $8n$ disk space for constructing an SA encoded by 40-bit integers. This indicates a great potential for improving space efficiency of the prior arts, such as DSA-IS and SAIS-PQ. In this paper, we present a checking method that enables any IS suffix sorting algorithms to check an SA when it is being built. For performance analysis, we integrate the proposed method into DSA-IS and implement the enhanced algorithm to evaluate the checking overhead, where the program design of DSA-IS has been optimized for a better speed and smaller peak disk use. From our experiments, the time, space and I/O consumptions for verification is negligible in comparison with that for construction. In view of the high performance of this checker and the existing IS suffix sorters, we are convinced that our checking method, in accompanied with the state-of-the-art IS suffix sorting algorithm, can constitute an efficient solution to the situations where checking is a must for a constructed SA.

@@

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Suffix array, construction and verification, external memory.
\end{IEEEkeywords}}

@@

\IEEEdisplaynontitleabstractindextext

\IEEEpeerreviewmaketitle

\section{Introduction}\label{sec:introduction}

Given a string drawn from a constant or integer alphabet, its suffix array can be built by the internal-memory algorithm SA-IS~\cite{Nong11} in linear time and space. So far, the IS method has been also applied to designing three external-memory suffix sorting algorithms~eSAIS~\cite{Bingmann12}, DSA-IS~\cite{Nong15} and SAIS-PQ~\cite{Liu15}, which have better time and I/O complexities than other alternatives, e.g., DC3~\cite{Dementiev2008a}, bwt-disk~\cite{Ferragina2012}, SAscan~\cite{Karkkainen2014} and pSAscan~\cite{Karkkainen2015}. However, the peak disk uses of their current programs are at least 15 times the sizes of input strings, resulting in a performance bottleneck. Recently, it was reported in~\cite{Karkkainen2017} a new IS suffix sorting algorithm fSAIS with a careful engineering that consumes nearly optimal disk space and runs faster than the parallel suffix sorter pSAscan on massive datasets. The algorithmic design of fSAIS is highly dependent on that of the prior arts. Particularly, it employs a sorting method similar to that of DSA-IS for inducing suffixes/substrings and the naming method of SAIS-PQ for producing reduced strings. The substantial improvement on space efficiency is mainly achieved by means of a monotone priority queue  for sorting substrings/suffixes and a garbage collector for quickly recycling disk space occupied by data no longer to use. As a result, its implementation takes less than $8n$ disk space for any input string of a size $n \le 2^{40}$, indicating a great potential for optimizing the programs for DSA-IS and SAIS-PQ. In this paper, we make our first attempt to redesign DSA-IS using new substring sorting and naming methods. The experimental results indicate that our program for the enhanced algorithm consumes about two-thirds as much disk space as that for eSAIS and runs faster than the latter on real world datasets. In addition, we also give a discussion on further improving our program to achieve a better space efficiency using the techniques for implementing fSAIS.

To detect potential computation errors caused by implementation bugs and/or hardware malfunctions, a constructed SA should be verified to ensure its correctness before use. This checking function has been included into various widespread software packages like SA-IS, DC3 and eSAIS. Particularly, the latter two verify an SA using the checking method presented in~\cite{Dementiev2008a}, which mainly consists of two passes of external-memory sorts involving $\mathcal{O}(n)$ fixed-size tuples with an integer key. In this paper, we propose a new checking method that enables any IS suffix sorting algorithm to check an SA when it is being built. 
For performance analysis, we integrate the proposed method into the optimized DSA-IS and implement the combined algorithm, called DSA-IS+, to evaluate the checking overhead. From our experiments, the time, space and I/O consumptions for verification is negligible in comparison with that for construction. In view of the high performance of this checker and the existing IS suffix sorters, we believe that our checking method, together with the state-of-the-art IS suffix sorting algorithm, can constitute an efficient solution to the situations where checking is a must for a constructed SA.

The rest of the paper is organized as follows. We first describe in Sections~\ref{sec:build_sa} and~\ref{sec:check_sa} the adapted algorithmic design of DSA-IS and the proposed checking method, respectively. Then, we demonstrate the experimental results in Section~\ref{sec:experiments} and draw the concluding remarks in Section~\ref{sec:conclusion}.

\section{SA Builder}~\label{sec:build_sa}

\subsection{Preliminaries}

Given an input string $x[0, n)$ with a unique smallest character $x[n - 1]$, we denote by ${\sf suf}(i)$ a suffix starting with $x[i]$ and ${\sf sub}(i,j)$ a substring running from $x[i]$ to $x[j]$. The following notations are used in our presentation for denotation convenience.

Character/substring/suffix classification. Characters of $x$ are classified into two categories: L-type and S-type. To be specific, $x[i]$ is S-type if (1) $i = n - 1$ or (2) $x[i] = x[i + 1]$ and $x[i + 1]$ is S-type; otherwise, $x[i]$ is L-type. Particularly, if $x[i]$ and $x[i - 1]$ are respectively S-type and L-type, then $x[i]$ is also S*-type. In addition, a substring/suffix is of the same type as its heading character. We use an integer array $t$ to indicate the type of characters, where $t[i] = 1$ or $0$ if $x[i] $ S-type or L-type, respectively.

Predecessor and Successor. For any two neighboring characters $x[i]$ and $x[i + 1]$, we say $x[i]$ is the predecessor of $x[i + 1]$ while the latter is the successor of the former. Similarly, there also exists a predecessor-successor relationship between ${\sf suf}(i)/{\sf sub}(i, j)$ and ${\sf suf}(i + 1)/{\sf sub}(i + 1, j)$.

\subsection{Overview of DSA-IS} \label{subsec:improvement}

% Algorithm
\begin{algorithm*}
	\SetAlgoNoLine
	\KwIn{$x$}
	\KwOut{$x_1$}

	$r := 0$ \label{alg1:1}

	\For{$i := n - 1; i \ge 0; i := i - 1$}{

		\If{$t[i] = 1$ and $t[i - 1] = 0$}{

			$PQ_1.{\sf push}(\langle x[i], t[i], r, i\rangle)$ // sort S*-type characters 
		}
	}\label{alg1:2}

	\While{$PQ_1.{\sf notEmpty}()$}{ \label{alg1:3}

		$top = PQ_1.{\sf top}(), PQ_1.{\sf pop()}, r:= r + 1$

		$pos = top.4th, ch = top.1st, pre\_ch = {\sf getPrecCh}(pos)$  // retrieve the heading character of the predecessor

		\If{$top.2nd = 0$ and $pre\_ch \ge ch$ or $top.2nd = 1$ and $pre\_ch > ch$}{

			$PQ_1.{\sf push}(\langle pre\_ch, 1, r, pos  - 1 \rangle)$ // induce the L-type predecessor
		}
		\Else{
			$PQ_2.{\sf push}(\langle ch, 1, r, pos \rangle)$ // keep the sorted L*-type substrings
		}
	} \label{alg1:4}

	$r := n$  \label{alg1:5}
	\While {$PQ_2.{\sf notEmpty}()$}{

		$top = PQ_2.{\sf top}(), PQ_2.{\sf pop()}, r := r - 1$

		$pos = top.4th, ch = top.1st, pre\_ch = {\sf getPrecCh}(pos)$  // retrieve the heading character of the predecessor

		\If{$top.2nd = 1$ and $pre\_ch \le ch$ or $top.2nd = 0$ and $pre\_ch < ch$}{

			$PQ_2.{\sf push}(\langle pre\_ch, 1, r, pos  - 1\rangle)$ // induce the S-type predecessor
		}
		\Else{

			$Vec.{\sf push\_back}(pos)$ // keep the sorted S*-type substrings
		}
	} \label{alg1:6}

	$last\_str := 0, last\_pos = n - 1, name = 0$ \label{alg1:7}

	\While {$Vec.{\sf notEmpty}()$}{

		$pos = Vec.{\sf pop\_back}(), str := {\sf getStr}(pos)$ // get S*-type substring

		\If{$last\_str \ne str$}{

			$name := name + 1$ // literally compare two lexicographically neighboring substrings 
		}

		$Sorter.{\sf push}(\langle pos, name\rangle)$

		$last\_pos = pos, last\_str = str$
	} \label{alg1:8}

	$Sorter.sort()$  \label{alg1:9}

	\For{$i := 0; Sorter.{\sf notEmpty}(); i:= i + 1$}{

		$x_1[i] = Sorter.2nd, Sorter.{\sf next}()$ // produce the reduced string

	} \label{alg1:10}

	\caption{The Reduction Phase for an IS Suffix Sorting Algorithm.}

	\label{alg:1}
\end{algorithm*}

An IS suffix sorting algorithm mainly consists of a reduction phase for sorting S*-type substrings and an induction phase for sorting suffixes. Both two phases execute the same process to accomplish the sorting tasks, while the reduction phase additionally performs a naming process to produce a reduced string $x_1$. For a better understanding, we illustrate in Algorithm~\ref{alg:1} the details of the reduction phase and summarize the framework below, where $PQ_1$ and $PQ_2$ are two priority queues with a sorting key consisting of the first three components.

\begin{enumerate}[S1]

\item Scan $x$ to find S*-type characters and insert them into $PQ_1$. (lines~\ref{alg1:1}-\ref{alg1:2}) \label{reduction_phase_framework:1}

\item Sequentially access the sorted substrings popped from $PQ_1$ in ascending order to induce their L-type predecessors, where each induced L-type substring is pushed into $PQ_1$ and each scanned L*-type substring is pushed into $PQ_2$ for later use. (lines~\ref{alg1:3}-\ref{alg1:4}) \label{reduction_phase_framework:2}

\item Sequentially access the sorted substrings popped from $PQ_2$ in descending order to induce their S-type predecessors, where each induced S-type substring is pushed into $PQ_2$ and each scanned S*-type substring is appended to a vector $Vec$. (lines~\ref{alg1:5}-\ref{alg1:6}) \label{reduction_phase_framework:3}

\item Compare the sorted S*-type substrings in $Vec$ to assign them names, where the name for each substring indicates it relative order among all. (lines~\ref{alg1:7}-\ref{alg1:8}) \label{reduction_phase_framework:4}

\item Sort the names in position order to produce the reduced string $x_1$. (lines~\ref{alg1:9}-\ref{alg1:10}) \label{reduction_phase_framework:5}

\end{enumerate}

It should be noticed that S\ref{reduction_phase_framework:2} and S\ref{reduction_phase_framework:3} induce the order of unsorted substrings from that of their heading characters and successors. When finished these two steps, all the S*-type substrings are sorted in their lexical order and stored in $Vec$. S\ref{reduction_phase_framework:4} performs a naming process to literally compare each pair of neighbors in the vector and assign them the same or different names with respect to whether they are equal or not. After that, S\ref{reduction_phase_framework:5} replaces each S*-type substring with its name to produce the reduced string $x_1$.

Algorithm~\ref{alg:1} can be implemented in external memory by sequential I/O operations except when calling ${\sf getPrecCh}$ or ${\sf getStr}$. These two functions invoke a random disk access per execution and thus may lead to a performance bottleneck. The existing disk-based IS suffix sorting algorithms adopt different approaches to amortize the I/O overhead. For DSA-IS, it employs an internal-memory preprocessor to prefetch the required data and sort them in their access order before use. In general, it first divides the input string into $\mathcal{O}(n/M)$ blocks consisting of one or multiple successive S*-type substrings. Then, it loads each block into RAM and sorts the S*-type substrings in the block following the same idea of S\ref{reduction_phase_framework:2}-S\ref{reduction_phase_framework:5}, where it records the returns of ${\sf getPrecCh}$ and ${\sf getStr}$ during the sort. Afterward, it executes Algorithm~\ref{alg:1} to sort all the S*-type substrings and retrieves the preceding characters and the target substrings from the block-wise results by sequential I/O operations.

\subsection{Improvements on Design of DSA-IS}

Recall that Algorithm~\ref{alg:1} performs the sorting and naming processes in sequence. Our first improvement is to reuse the substring naming method originally proposed for SAIS-PQ to combine the execution of these two processes during the reduction phase. In Algorithm~\ref{alg:1}, two priority queues are employed to sort substrings according to their heading characters and the rank of their successors, where the rank value indicates the access order of the corresponding substring. Notice that the lexical order of any two substrings/suffixes can be determined by that of their heading characters and successors as well, thus the priority queues can sort substrings by their names instead of their ranks. The name of currently scanned L-type/S-type substring can be computed by comparing its heading character and the name of its successor with that of the last scanned L-type/S-type substring. If both are equal, then the two substrings have the same name; otherwise, the former is assigned a name bigger/smaller than the latter.

From our experiments, sorting substrings by the priority queues takes long CPU time and large I/O volume. To reduce the consumption, we present a new method that combines the sorted S*-type substrings of all the blocks by a merger, which maintains a buffer for each block to preload the sorted S*-type substrings into RAM and an internal-memory heap to compare the substrings belonging to different blocks by literal comparisons. For the substrings too long to be wholly loaded into RAM, this method first uses the priority queue to sort them in external memory and then employs the merger to compare them with the short substrings based on the following property: suppose $sub_a$ and $sub_b$ are respectively long and short substrings, then $sub_a < sub_b$ if (1) $sub_a[0, D) = sub_b$ or (2) $sub_a[0, k) = sub_b[0, k)$ and $sub_af_a[k] < sub_b[k]$, where $D = \|sub_b\|$ and $k < D$. This indicates that we only need to fetch the first $D$ characters of a long substring into RAM when comparing it with a short substring containing $D$ characters.

\subsection{Discussion} \label{subsec:further_discussion}

As demonstrated in Section~\ref{sec:experiments}, our implementation for the enhanced DSA-IS has an advantage over that for eSAIS in terms of time, space and I/O volume, but its peak disk use remains at a high level. This is mainly because it fails to gradually clean up the temporary data when they are being used. On the contrary, fSAIS stores the temporary data using multiple files and deletes each file immediately when all that residing on it is longer to use. Therefore, it is reasonable to believe that our program can achieve a substantial improvement following the same technique. Currently, we are attempting to further improve the algorithmic and implementation designs of DSA-IS.

\section{SA Checker}~\label{sec:check_sa}

\subsection{Details}

A constructed SA should be verified to ensure its correctness before use. This checking function can be found in some widespread software packages like SA-IS, DC3 and eSAIS, where the latter two perform verification based on Theorem~\ref{theorem:1}.

\begin{theorem} \label{theorem:1}
	$sa[0, n)$ is the SA for $x[0, n)$ if and only if the following conditions are satisfied:\\
	(1) $sa$ is a permutation of $[0, n)$. \\	
	(2) ${\sf r_i < r_j} \Leftrightarrow (x[i], r_{i + 1}) < (x[i], r_{j + 1})$ for $ i, j \in [0, n)$ and $i\ne j$, where $r_i$ and $r_j$ indicates the relative order of ${\sf suf}(i)$ and ${\sf suf}(j)$ among all the suffixes, respectively. \\
\end{theorem}

\begin{IEEEproof}
	The proof is available in~\cite{Dementiev2008a}.
\end{IEEEproof}

This method executes two passes of integer sorts and each sort involves $n$ fixed-size tuples, thus it takes large disk space and I/O volume when implemented in the external memory, resulting in a non-negligible checking overhead. Take eSAIS for example, the peak disk use for sequentially building and checking SA rises up to $27n$, while the I/O volume also increases by $50n$.
